TrenchR: an R package for transparent environmental and ecological biophysics

### 2022-06-28 Manuscript_full.Rmd

```{r, echo=FALSE, fig.height=4, out.width = "75%"}
knitr::include_graphics('./figures/TrenchRBiophys.png')
```

@@

**Utility functions** aid calculation of environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for oranismal development.

**Allometric functions** allow estimating the dimensions of organisms needed for energy balances and other analyses. The functions can convert between organismal length, mass, surface area, and volume. The functions can also estimate the silhoutte area, which describes the organismal area exposed to solar radiation.

@@

**General biophysical functions** provide generalized models of heat exchanges between organisms and their environment so that users can build custom biophysical models. 
The functions allow implementing an energy balance including the following components of heat exchange between organisms and their environment: 
  *radiative heat exchanges* of solar and thermal radiation; 
  *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; 
  *conductive heat exchanges* between organisms and solid surfaces (generally the ground) due to physical contact; and 
  *evaporative heat exchanges* associated with organisms' evaporative water loss. 
Additional functions aggregate these forms of heat exchange into energy balances and to use the energy balances to predict body temperatures.

@@

```{r, echo=FALSE, fig.height=4, out.width = "75%"}
knitr::include_graphics('./figures/TrenchRFlow.png')
```

@@

We introduce the functions in several vignettes. 
A good place to start is the *Allometry and conversions* vignette, which provides tools for preparing data, such as estimating additional dimensions of organisms from measured dimensions. The *Estimating microclimates* vignette provides resources for estimating the environmental conditions experienced by organisms. 
This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and wind speed profiles, and soil temperatures and profiles.
Finally, the core biophysical modeling functions are described in a tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models. 
We additionally offer a *List of symbols* used in equations.

@@

```{r}
# Set up input data as variables
lat = 35.69; # Latitude (degrees)
lon = -105.944; # Longitude (degrees)
elev = 2121 # Elevation (meters)

Tmin = 10; # Minimum air temperature (C)
Tmax = 25; # Maximum air temperature (C)
Tmin_s = 15; # Minimum soil temperature (C)
Tmax_s = 30; # Maximum soil temperature (C)
V = 1 # Wind speed (m/s)

# Assumptions
tau = 0.7; # Atmospheric transmissivity
rho = 0.6; # Albedo
Tb0 = 25; # Initial assumption of body temperature (C)

doy = day_of_year("2021-06-01", format= "%Y-%m-%d") # Find Julian Date (DOY)
snoon = solar_noon(lon = lon, doy = doy) # Estimate solar noon
dayl = daylength(lat = lat, doy = doy) # Estimate day length
tr = snoon - dayl / 2 # Time of sunrise
ts = snoon + dayl / 2 # Time of sunset
```

@@

```{r}
# Estimate zenith angle (degrees)
psi_deg = sapply(1:24, FUN = zenith_angle, doy = doy, lat = lat, lon = lon) 
# Convert to radians
psi_rad = degrees_to_radians(psi_deg) 
# Estimate radiation
Srad = sapply(psi_rad, FUN = solar_radiation, doy = doy, tau = tau, elev = elev, rho = rho)

# Separate solar radiation into direct, diffuse, and reflected
Sdir = Srad[1,]; # Direct solar radiation
Sdif = Srad[2,]; # Diffuse solar radiation
Sref = Srad[3,]; # Reflected solar radiation
```

@@

```{r}
# Air temperature ($^\circ$C)
Ta = sapply(1:24, diurnal_temp_variation_sineexp, T_max = Tmax, T_min = Tmin, t_r = tr, t_s = ts, alpha = 2.59, beta = 1.55, gamma = 2.2) 
# Soil temperature ($^\circ$C)
Ts = sapply(1:24, diurnal_temp_variation_sine, T_max = Tmax_s, T_min = Tmin_s) 
```

@@

```{r}
# Scale air temperature to lizard height
# T_r = reference temperature; zr = reference height; z0 is surface roughness; z = organism height; T_s = surface temperature
Ta_liz = air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Neutral air temperature profile
Ta_liz2 = air_temp_profile(T_r = Ta, u_r = V, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Air temperature profile forced by wind speed

# Scale wind speed to lizard height
# u_r = reference wind speed; zr = reference height; z0 is surface roughness; z = organism height
V_liz= wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) # Neutral wind speed profile
```

@@

```{r}
mass = 10 # Mass (g)
a = 0.9 # Solar absorptivity (proportion)

# Estimate surface area and the proportion sihouette area
A = surface_area_from_mass(mass, "lizard") # surface area (m^2)
psa = sapply(psi_deg, proportion_silhouette_area, taxon = "lizard", posture = "prostrate") 
# Change negative values to zero
psa[psa < 0] = 0
```

@@

```{r}
Qabs = rep(NA, 24)
for(hour in 1:24) {

  Qabs[hour] = Qradiation_absorbed(a = a, A = A, psa_dir = psa[hour], psa_ref = 1 - psa[hour], S_dir = Sdir[hour], S_dif = Sdif[hour], a_s = rho)

@@

```{r}
epsilon_s = 0.965 # Surface emissivity of lizards

Qemit = rep(NA, 24)
for(hour in 1:24) {

  # psa_dir = proportions of surface area exposed to direct solar radiation
  # psi_ref = proportions of surface area exposed to reflected solar radiation
  # A = surface area (m^2)
  Qemit[hour] = Qemitted_thermal_radiation(epsilon = epsilon_s, A = A, psa_dir = psa[hour], psa_ref = 1-psa[hour], T_b = Ta_liz[hour] + 273.15, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, enclosed = FALSE)

@@

```{r}
ap = airpressure_from_elev(elev) * 1000 # Barometric pressure (pascal)

# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
DRYAIRout = DRYAIR(db = Ta, bp = ap, alt = elev)
K = DRYAIRout$thcond # Thermal conductivity (Wm^-2K^-1) 
nu = DRYAIRout$viskin # Kinematic viscosity (m2 s-1)

svl = 0.006 # Approximate snout vent length (meters) for Sceloporus, used as the characteristic dimension for heat transfer

# We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.
K = mean(K) # Average thermal conductivity across the day
nu = mean(nu) # Average kinematic viscosity

# Estimate the heat transfer coefficient using an empirical relationship for lizards
H_L = heat_transfer_coefficient(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard_surface")

# Estimate the heat transfer coefficient using a spherical approximation
H_L2= heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard")

# Estimate the heat transfer coefficient using a simplified version of the approximation
H_L3= heat_transfer_coefficient_simple(V = V_liz, D = svl, type = "Gates")
```

@@

```{r}
Qconv = rep(NA, 24)
for(hour in 1:24) {

  Qconv[hour] = Qconvection(T_a = Ta_liz[hour] + 273.15, T_b = Ta_liz[hour] + 273.15, H = H_L, A = A, proportion = 0.67, ef = 1.3)

@@

```{r}
Qcond = rep(NA, 24)
for(hr in 1:24) {

  Qcond[hr] = Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 10^-6, K = K, A = A, proportion = 0.33)

@@

```{r}
Qmet = 0; 
Qevap= 0;
```

@@

```{r}
Qnet = Qnet_Gates(Qabs = Qabs, Qemit = Qemit, Qconv = Qconv, Qcond = Qcond, Qmet = Qmet, Qevap = Qevap)
```

@@

```{r}
Te = rep(NA, 24)
for(hour in 1:24) {

  Te[hour]= Tb_Gates(A = A, D = svl, psa_dir = psa[hour], psa_ref = 1 - psa[hour], psa_air = 0.67, psa_g = 0.33, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K)

@@

```{r}
Te2 = rep(NA, 24)
for(hr in 1:24) {

  # S is solar radiation flux (W m^-2), so we divide by surface area, A
  Te2[hr] = Tb_CampbellNorman(T_a = Ta_liz[hr] + 273.15, T_g = Ts[hr] + 273.15, S = Qabs[hr] / A, alpha_L = 0.96, epsilon = epsilon_s, c_p = 29.3, D = svl, V = V_liz)

@@

```{r}
Te3 = rep(NA, 24)
for(hour in 1:24) {

  # F_d, F_r, F_a, and F_g are the view factors between the surface of the lizard and diffuse solar radiation, reflected solar radiation, and atmospheric thermal radiation, and ground thermal radiation, respectively. 
  Te3[hour]= Tb_lizard(T_a = Ta_liz[hour], T_g = Ts[hour], u = V_liz, svl = svl * 1000, m = mass, psi = psi_deg[hour], rho_S = rho, elev = elev, doy = doy, sun = TRUE, surface = TRUE, alpha_S = a, alpha_L = 0.965, epsilon_s = epsilon_s, F_d = 0.8, F_r = 0.5, F_a = 0.5, F_g = 0.5)

@@

```{r, echo=FALSE}
par(mar=c(5, 4, 4, 8) + 0.1)
plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab= expression("Temperature ("*~degree*C*")"), col="#5DC863FF", ylim=c(10,60)) # Gates
points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") # Campbell and Norman
points(1:24, Te3, type="l", col="#5DC863FF", lty="dashed") # Lizard, Buckley 2008
points(1:24, Ta, type="l", col="#440154FF")
points(1:24, Ta_liz, type="l", col="#3B528BFF")
points(1:24, Ts, type="l", col="#21908CFF")

#add additional axis with radiation
par(new = T)
plot(1:24, Sdir, pch=16, axes=F, xlab=NA, ylab=NA, type="l", col="#FDE725FF")
axis(side = 4)
mtext(side = 4, line = 3, 'Radiation (W/m^2)')
legend("topleft",
       legend=c("Ta (2m)", "Ta (0.02m)", "Ts", "Te", "Radiation"),
       lty=1, pch=NA, col=c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF"))
```

>>>>

```{r, echo=FALSE, fig.height=4, fig.align = 'center', out.width = "75%"}
knitr::include_graphics('./figures/TrenchRBiophys.png')
```

@@

**Utility functions** calculate environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for oranismal development.

**Allometric functions** allow estimating the dimensions of organisms needed for energy balances and other analyses. 
Available functions can convert between organismal length, mass, surface area, and volume as well as estimate the silhoutte area, which describes the organismal area exposed to solar radiation.

@@

**General biophysical functions** provide generalized models of heat exchanges between organisms and their environment so that users can build custom biophysical models. 
The functions allow implementing an energy balance including the following components of heat exchange between organisms and their environment:

  - *radiative heat exchanges* of solar and thermal radiation; 
  - *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; 
  - *conductive heat exchanges* between organisms and solid surfaces (generally the ground) due to physical contact; and 
  - *evaporative heat exchanges* associated with organisms' evaporative water loss.

Additional functions aggregate these forms of heat exchange into energy balances and to use the energy balances to predict body temperatures.

@@

```{r, echo=FALSE, fig.height=4, fig.align = 'center', out.width = "75%"}
knitr::include_graphics('./figures/TrenchRFlow.png')
```

@@

We introduce the functions in several vignettes. 
A good place to start is the *Allometry and Conversions* vignette, which provides tools for preparing data, such as estimating additional dimensions of organisms from measured dimensions. The *Estimating Microclimates* vignette provides resources for estimating the environmental conditions experienced by organisms. 
This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and wind speed profiles, and soil temperatures and profiles.
Finally, the core biophysical modeling functions are described in a tutorial on *Using Energy Balances to Estimate Body Temperatures*.
Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models. 
We additionally offer a *List of Symbols* used in equations.

@@

```{r}
# Set up input data as variables

lat  <- 35.69    # Latitude (degrees)
lon  <- -105.944 # Longitude (degrees)
elev <- 2121     # Elevation (meters)

Tmin   <- 10 # Minimum air temperature (C)
Tmax   <- 25 # Maximum air temperature (C)
Tmin_s <- 15 # Minimum soil temperature (C)
Tmax_s <- 30 # Maximum soil temperature (C)
V      <- 1  # Wind speed (m/s)

# Assumptions

tau <- 0.7 # Atmospheric transmissivity
rho <- 0.6 # Albedo
Tb0 <- 25  # Initial assumption of body temperature (C)

doy   <- day_of_year("2021-06-01", format = "%Y-%m-%d") # Julian Date (DOY)

snoon <- solar_noon(lon = lon, doy = doy) # Estimate solar noon
dayl  <- daylength(lat = lat, doy = doy)  # Estimate day length

tr    <- snoon - dayl / 2 # Time of sunrise
ts    <- snoon + dayl / 2 # Time of sunset
```

@@

```{r}
# Estimate zenith angle (degrees)

psi_deg <- sapply(1:24, FUN = zenith_angle, doy = doy, lat = lat, lon = lon)

# Convert to radians

psi_rad <- degrees_to_radians(psi_deg)

# Estimate radiation

Srad <- sapply(psi_rad, FUN = solar_radiation, doy = doy, tau = tau, elev = elev, rho = rho)

# Separate solar radiation into direct, diffuse, and reflected

Sdir <- Srad[1,] # Direct solar radiation
Sdif <- Srad[2,] # Diffuse solar radiation
Sref <- Srad[3,] # Reflected solar radiation
```

@@

```{r}
# Air temperature (C)

Ta <- sapply(1:24, diurnal_temp_variation_sineexp, T_max = Tmax, T_min = Tmin, t_r = tr, t_s = ts, alpha = 2.59, beta = 1.55, gamma = 2.2)

# Soil temperature (C)

Ts <- sapply(1:24, diurnal_temp_variation_sine, T_max = Tmax_s, T_min = Tmin_s) 
```

@@

```{r}
# Scale air temperature to lizard height
# T_r = reference temperature
# zr  = reference height
# z0  = surface roughness
# z   = organism height
# T_s = surface temperature

Ta_liz  <- air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts)  # Neutral air temperature profile
Ta_liz2 <- air_temp_profile(T_r = Ta, u_r = V, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Air temperature profile forced by wind speed

# Scale wind speed to lizard height
# u_r = reference wind speed
# zr  = reference height
# z0  = surface roughness
# z   = organism height

V_liz <- wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) # Neutral wind speed profile
```

@@

```{r}
mass <- 10  # Mass (g)
a    <- 0.9 # Solar absorptivity (proportion)

# Estimate surface area (m^2) and the proportion sihouette area

A   <- surface_area_from_mass(mass, "lizard")  
psa <- sapply(psi_deg, proportion_silhouette_area, taxon = "lizard", posture = "prostrate")

# Change negative values to zero
psa[psa < 0] = 0
```

@@

```{r}
Qabs <- rep(NA, 24)
for (hour in 1:24) {

  Qabs[hour] <- Qradiation_absorbed(a = a, A = A, psa_dir = psa[hour], psa_ref = 1 - psa[hour], S_dir = Sdir[hour], S_dif = Sdif[hour], a_s = rho)

@@

```{r}
epsilon_s <- 0.965 # Surface emissivity of lizards

Qemit <- rep(NA, 24)
for (hour in 1:24) {

  # psa_dir = proportions of surface area exposed to direct solar radiation
  # psi_ref = proportions of surface area exposed to reflected solar radiation
  # A       = surface area (m^2)

  Qemit[hour] <- Qemitted_thermal_radiation(epsilon = epsilon_s, A = A, psa_dir = psa[hour], psa_ref = 1-psa[hour], T_b = Ta_liz[hour] + 273.15, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, enclosed = FALSE)

@@

```{r}
ap <- airpressure_from_elev(elev) * 1000 # Barometric pressure (pascal)

# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity

DRYAIRout <- DRYAIR(db = Ta, bp = ap, alt = elev)

K  <- DRYAIRout$thcond # Thermal conductivity (Wm^-2K^-1) 
nu <- DRYAIRout$viskin # Kinematic viscosity (m2 s-1)

svl <- 0.006 # Approximate snout vent length (meters) for Sceloporus, used as the characteristic dimension for heat transfer

# We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.

K  <- mean(K)  # Average thermal conductivity across the day
nu <- mean(nu) # Average kinematic viscosity

# Estimate the heat transfer coefficient using an empirical relationship for lizards

H_L <- heat_transfer_coefficient(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard_surface")

# Estimate the heat transfer coefficient using a spherical approximation

H_L2 <- heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard")

# Estimate the heat transfer coefficient using a simplified version of the approximation

H_L3 <- heat_transfer_coefficient_simple(V = V_liz, D = svl, type = "Gates")
```

@@

```{r}
Qconv <- rep(NA, 24)
for (hour in 1:24) {

  Qconv[hour] <- Qconvection(T_a = Ta_liz[hour] + 273.15, T_b = Ta_liz[hour] + 273.15, H = H_L, A = A, proportion = 0.67, ef = 1.3)

@@

```{r}
Qcond <- rep(NA, 24)
for(hr in 1:24) {

  Qcond[hr] <- Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 10^-6, K = K, A = A, proportion = 0.33)

@@

```{r}
Qmet  <- 0 
Qevap <- 0
```

@@

```{r}
Qnet <- Qnet_Gates(Qabs = Qabs, Qemit = Qemit, Qconv = Qconv, Qcond = Qcond, Qmet = Qmet, Qevap = Qevap)
```

@@

```{r}
Te <- rep(NA, 24)
for (hour in 1:24) {

  Te[hour] <- Tb_Gates(A = A, D = svl, psa_dir = psa[hour], psa_ref = 1 - psa[hour], psa_air = 0.67, psa_g = 0.33, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K)

@@

```{r}
Te2 <- rep(NA, 24)
for (hr in 1:24) {

  # S is solar radiation flux (W m^-2), so we divide by surface area, A

  Te2[hr] <- Tb_CampbellNorman(T_a = Ta_liz[hr] + 273.15, T_g = Ts[hr] + 273.15, S = Qabs[hr] / A, alpha_L = 0.96, epsilon = epsilon_s, c_p = 29.3, D = svl, V = V_liz)

@@

```{r}
Te3 <- rep(NA, 24)
for (hour in 1:24) {

  # F_d, F_r, F_a, and F_g are the view factors between the surface of the lizard and diffuse solar radiation, reflected solar radiation, and atmospheric thermal radiation, and ground thermal radiation, respectively.

  Te3[hour] <- Tb_lizard(T_a = Ta_liz[hour], T_g = Ts[hour], u = V_liz, svl = svl * 1000, m = mass, psi = psi_deg[hour], rho_S = rho, elev = elev, doy = doy, sun = TRUE, surface = TRUE, alpha_S = a, alpha_L = 0.965, epsilon_s = epsilon_s, F_d = 0.8, F_r = 0.5, F_a = 0.5, F_g = 0.5)

@@

```{r, echo=FALSE}
par(mar = c(5, 4, 4, 8) + 0.1)

plot(1:24, Te - 273.15, type = "l", xlab = "Hour", ylab = expression("Temperature ("*~degree*C*")"), col = "#5DC863FF", ylim = c(10,60), las = 1) # Gates
points(1:24, Te2 - 273.15, type ="l", col = "#5DC863FF", lty = "dotted") # Campbell and Norman
points(1:24, Te3, type = "l", col = "#5DC863FF", lty = "dashed") # Lizard, Buckley 2008
points(1:24, Ta, type = "l", col = "#440154FF")
points(1:24, Ta_liz, type = "l", col = "#3B528BFF")
points(1:24, Ts, type = "l", col = "#21908CFF")

# add additional axis with radiation

par(new = TRUE)
plot(1:24, Sdir, pch = 16, axes = FALSE, xlab = NA, ylab = NA, type = "l", col = "#FDE725FF")
axis(side = 4, las = 1)
mtext(side = 4, line = 3, 'Radiation (W/m^2)')
legend("topleft",
       legend = c("Ta (2m)", "Ta (0.02m)", "Ts", "Te", "Radiation"),
       lty = 1, pch = NA, col = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF"))
```
